---
title: "Final Project, Winter 2021"
author: "Alex Liddle"
date: "3/2/2021"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=TRUE}

# Load any packages, if any, that you use as part of your answers here
# For example: 
library(plyr)
library(data.table)
library(ggplot2)
#install.packages('GGally')
library(GGally)
library(tidyverse)
library(ResourceSelection)
library(glmnet)
#install.packages('mvnormtest')
library(mvnormtest)
#install.packages('sgd')
library(sgd)
#install.packages("factoextra")
library(factoextra)
```

CONTEXT -- 

Given the incredible size of many datasets, is it valid approach to use Stochastic Gradient Descent (SGD) for approximating the cost function when fitting a statistical model?

SGD significantly reduces the computational cost of approximating a cost function by looking at a small random sample, versus the entire dataset. 

The data source contains airline departure delay information from the department of transportation from 2008 to 2019 and was compiled into a single dataset on kaggle.com:
https://www.kaggle.com/sherrytp/airline-delay-analysis/. There are 26 variables, although some of them won’t be used (Note, the purpose of the statistical model will be to predict delay times, so variables such as total_flight_time, wheels_on, taxi_in, etc that explicitly describe the arrival delay will not be used). The variables in the dataset are as follows:

FL_DATE	-- Flight Date (yyyy-mm-dd)	 
OP_CARRIER -- Airline
OP_CARRIER_FL_NUM -- Airline	Flight Number	 
ORIGIN --	Origin Airport, Airport ID. An identification number assigned by US DOT to identify a unique airport. Do not use this field for airport analysis across a range of years because an airport can change its airport code and airport codes can be reused.
DEST --	Destination Airport, Airport ID. An identification number assigned by US DOT to identify a unique airport. Do not use this field for airport analysis across a range of years because an airport can change its airport code and airport codes can be reused.
CRS_DEP_TIME -- CRS Departure Time, e.g. scheduled departure time (local time: hhmm)	 
DEP_TIME --	Actual Departure Time (local time: hhmm)	
DEP_DELAY -- Difference in minutes between scheduled and actual departure time. Early departures show negative numbers.
TAXI_OUT --	Taxi Out Time, in Minutes
WHEELS_OFF --	Wheels Off Time (local time: hhmm)	 
WHEELS_ON --	Wheels On Time (local time: hhmm)	 
TAXI_IN --	Taxi In Time, in Minutes
CRS_ARR_TIME --	CRS Arrival Time (local time: hhmm)	 
ARR_TIME --	Actual Arrival Time (local time: hhmm)	 
ARR_DELAY --	Difference in minutes between scheduled and actual arrival time. Early arrivals show negative numbers.
CANCELLED --	Cancelled Flight Indicator (1=Yes)
CANCELLATION_CODE --	Specifies The Reason For Cancellation
DIVERTED --	Diverted Flight Indicator (1=Yes)
CRS_ELAPSED_TIME --	CRS Elapsed Time of Flight, in Minutes
ACTUAL_ELAPSED_TIME --	Elapsed Time of Flight, in Minutes
AIR_TIME -- Elapsed Time in Air, in Minutes
DISTANCE -- Distance Traveled, in Miles
CARRIER_DELAY --	Carrier Delay, in Minutes	Analysis
WEATHER_DELAY --	Weather Delay, in Minutes	Analysis
NAS_DELAY --	National Air System Delay, in Minutes
SECURITY_DELAY --	Security Delay, in Minutes
LATE_AIRCRAFT_DELAY --	Late Aircraft Delay, in Minutes

Let us begin!

Load the data

```{r}

file_names <- list.files(path = "./",pattern = ".csv")

delays <- do.call('rbind.fill',lapply(file_names,read.csv))

delays <- as.data.table(delays)

delays <- delays[,which(unlist(lapply(delays, function(x)!all(is.na(x))))),with=F]

delays <- as.data.frame(delays)

delays <- delays[complete.cases(delays), ]

str(delays)
```

Assumptions of data:
"Kassambara, &amp; U, M. (2018, March 11). Logistic regression assumptions and diagnostics in r. Retrieved March 7, 2021, from http://www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/#linearity-assumption"

- binary dependent variable
- independent observations
- little or no multicollinearity among the independent variables

Data Preprocessing

```{r}
# Remove canceled and diverted flights since we are only interested in ones that have been delayed
delays$CANCELLED <- as.factor(delays$CANCELLED)
delays$DIVERTED <- as.factor(delays$DIVERTED)
delays <- delays[!(delays$CANCELLED=="1" | delays$DIVERTED=="1"),]

# Remove unwanted columns (Because the goal is to predict arrival delay, there is no need for data related to the arrival, except for the label we are trying to predict (ARR_DELAY))
delays <- delays[ , -which(names(delays) %in% c("OP_CARRIER","OP_CARRIER_FL_NUM", "CANCELLED", "CANCELLATION_CODE","DIVERTED", "ORIGIN", "DEST", "CRS_ARR_TIME", "CRS_ELAPSED_TIME", "ACTUAL_ELAPSED_TIME", "TAXI_IN", "WHEELS_ON", "ARR_TIME"))]
```

```{r}
# Convert to numerical data types
delays$FL_DATE <- as.numeric(as.Date(delays$FL_DATE))
delays$CRS_DEP_TIME <- as.numeric(delays$CRS_DEP_TIME)
delays$DEP_DELAY <- as.numeric(delays$DEP_DELAY)
delays$DEP_TIME <- as.numeric(delays$DEP_TIME)
delays$TAXI_OUT <- as.numeric(delays$TAXI_OUT)
delays$WHEELS_OFF <- as.numeric(delays$WHEELS_OFF)
delays$ARR_DELAY <- as.numeric(delays$ARR_DELAY)
```


```{r}
# Create binomial output variable: 1 for a long delay, 0 for a short delay
delays$ARR_DELAY <- ifelse(delays$ARR_DELAY > quantile(as.matrix(delays$ARR_DELAY), 0.5), 1, 0)

str(delays)
```

Test assumptions of data

```{r}
set.seed(42)

# Test for collinearity and remove columns if necessary
cor(as.matrix(delays[ , -which(names(delays) %in% c("ARR_DELAY"))]))
ggpairs(delays[ , -which(names(delays) %in% c("ARR_DELAY"))][sample(nrow(delays), 1000), ])

# Conduct PCA to remove high correlation without losing feature data
res.pca <- prcomp(delays[ , -which(names(delays) %in% c("ARR_DELAY"))], scale = TRUE)
eig.val <- get_eigenvalue(res.pca)

# Get components that explain more than 10% of variance
num_components <- eig.val$variance.percent[eig.val$variance.percent >= 10]

delays.pca <- predict(res.pca, newdata = delays[ , -which(names(delays) %in% c("ARR_DELAY"))])

delays.pca <- data.frame(delays.pca[,1:length(num_components)])

cor(as.matrix(delays.pca))
ggpairs(delays.pca[sample(nrow(delays.pca), 1000), ])

```

SGD
"Tran, D., Toulis, P., &amp; Airoldi, E. M. (n.d.). Stochastic Gradient Descent Methods for Estimation with Large Data Sets. Journal of Statistical Software."

(great for exceptionally large datasets)

Build model

```{r}
delays.pca$ARR_DELAY <- delays$ARR_DELAY

n<-nrow(delays.pca)
set.seed(42)

train<-sample(rep(0:1,c(round(n*.3),n-round(n*.3))),n)

delays.train <- delays.pca[train==1,]
delays.test <- delays.pca[train==0,]

X = as.matrix(delays.train[ , -which(names(delays.train) %in% c("ARR_DELAY"))])
y = as.matrix(delays.train$ARR_DELAY)

m.sgd <- sgd(y~X[,1]+X[,2]+X[,3]+X[,4]+X[,5], model="glm", model.control=list(family="binomial"))
str(m.sgd$coefficients)
```

Now compare to the model built by the glm method

```{r}
nam<-names(delays.train)[1:(ncol(delays.train)-1)]

# Use stringr to avoid typing all the explanatory variables.
fmla<-as.formula(str_c("ARR_DELAY~",
    str_c(nam,collapse="+")))

print(fmla)

m<-glm(fmla,data=delays.train,family="binomial")

summary(m)
```

There are slight differences in the estimated coefficients. Notably, the time it takes to estimate the coefficients is significantly less for the SGD model.

Model Evaluation

## Confusion matrix

```{r}
# m.sgd
print("Confusion Matrix:")
print("m.sgd")
probs<-predict(m.sgd, as.matrix(delays.train), type="response")
pred<-probs>=.5
(confus.mat.m.sgd<-table(delays.train$ARR_DELAY,pred))

probs.test<-predict(m.sgd, as.matrix(delays.test), type="response")
pred.test<-probs.test>=.5
(confus.mat.test.m.sgd<-table(delays.test$ARR_DELAY,pred.test))

# glm
print("m.glm")
probs<-predict(m, delays.train[ , -which(names(delays.train) %in% c("ARR_DELAY"))], type="response")
pred<-probs>=.5
(confus.mat.m.glm<-table(delays.train$ARR_DELAY,pred))

probs.test<-predict(m, delays.test[ , -which(names(delays.train) %in% c("ARR_DELAY"))], type="response")
pred.test<-probs.test>=.5
(confus.mat.test.m.glm<-table(delays.test$ARR_DELAY,pred.test))
```

## Accuracy

Accuracy is the proportion correct.

```{r}
# m.sgd
print("Accuracy:")
print("m.sgd")
print("Train")
(acc.train.m.sgd<-sum(diag(confus.mat.m.sgd))/sum(confus.mat.m.sgd))
print("Test")
(acc.test.m.sgd<-sum(diag(confus.mat.test.m.sgd))/sum(confus.mat.test.m.sgd))

# m.glm
print("m.glm")
print("Train")
(acc.train.m.glm<-sum(diag(confus.mat.m.glm))/sum(confus.mat.m.glm))
print("Test")
(acc.test.m.glm<-sum(diag(confus.mat.test.m.glm))/sum(confus.mat.test.m.glm))
```

## Precision

Precision is the proportion of true positives to positives.

```{r}
# m.sgd
print("Precision:")
print("m.sgd")
print("Train")
(prec.train.m.sgd<-confus.mat.m.sgd[2,2]/sum(confus.mat.m.sgd[,2]))
print("Test")
(prec.test.m.sgd<-confus.mat.test.m.sgd[2,2]/sum(confus.mat.test.m.sgd[,2]))

# m.glm
print("m.glm")
print("Train")
(prec.train.m.glm<-confus.mat.m.glm[2,2]/sum(confus.mat.m.glm[,2]))
print("Test")
(prec.test.m.glm<-confus.mat.test.m.glm[2,2]/sum(confus.mat.test.m.glm[,2]))
```

## Recall

Recall is the proportion of true positives to all values of "1".

```{r}
# m.sgd
print("Recall:")
print("sgd")
print("Train")
(recall.train.m.sgd<-confus.mat.m.sgd[2,2]/sum(confus.mat.m.sgd[2,]))
print("Test")
(recall.test.m.sgd<-confus.mat.test.m.sgd[2,2]/sum(confus.mat.test.m.sgd[2,]))

# m.glm
print("glm")
print("Train")
(recall.train.m.glm<-confus.mat.m.glm[2,2]/sum(confus.mat.m.glm[2,]))
print("Test")
(recall.test.m.glm<-confus.mat.test.m.glm[2,2]/sum(confus.mat.test.m.glm[2,]))
```

## F1

F1 Score = 2(Recall)(Precision)/ (Recall + Precision)

```{r}
# m.sgd
print("F1-Score:")
print("m.sgd")
print("Train")
(F1.train.m.sgd<-2*recall.train.m.sgd*prec.train.m.sgd/(recall.train.m.sgd+prec.train.m.sgd))
print("Test")
(F1.test.m.sgd<-2*recall.test.m.sgd*prec.test.m.sgd/(recall.test.m.sgd+prec.test.m.sgd))

# m.glm
print("m.glm")
print("Train")
(F1.train.m.glm<-2*recall.train.m.glm*prec.train.m.glm/(recall.train.m.glm+prec.train.m.glm))
print("Test")
(F1.test.m.glm<-2*recall.test.m.glm*prec.test.m.glm/(recall.test.m.glm+prec.test.m.glm))
```


Conclusion:

Stochastic Gradient Descent is an incredibly powerful tool for cutting down the computing time when fitting exceptionally large datasets to statistical models.