---
title: "Final Project, Winter 2021"
author: "Alex Liddle"
date: "3/2/2021"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=TRUE}

# Load any packages, if any, that you use as part of your answers here
# For example: 
library(plyr)
library(data.table)
library(ggplot2)
library(tidyverse)
library(glmnet)
```

CONTEXT -- TODO (Describe research question and dataset)

FL_DATE	-- Flight Date (yyyy-mm-dd)	 
OP_CARRIER -- Airline
OP_CARRIER_FL_NUM -- Airline	Flight Number	 
ORIGIN --	Origin Airport, Airport ID. An identification number assigned by US DOT to identify a unique airport. Do not use this field for airport analysis across a range of years because an airport can change its airport code and airport codes can be reused.
DEST --	Destination Airport, Airport ID. An identification number assigned by US DOT to identify a unique airport. Do not use this field for airport analysis across a range of years because an airport can change its airport code and airport codes can be reused.
CRS_DEP_TIME -- CRS Departure Time, e.g. scheduled departure time (local time: hhmm)	 
DEP_TIME --	Actual Departure Time (local time: hhmm)	
DEP_DELAY -- Difference in minutes between scheduled and actual departure time. Early departures show negative numbers.
TAXI_OUT --	Taxi Out Time, in Minutes
WHEELS_OFF --	Wheels Off Time (local time: hhmm)	 
WHEELS_ON --	Wheels On Time (local time: hhmm)	 
TAXI_IN --	Taxi In Time, in Minutes
CRS_ARR_TIME --	CRS Arrival Time (local time: hhmm)	 
ARR_TIME --	Actual Arrival Time (local time: hhmm)	 
ARR_DELAY --	Difference in minutes between scheduled and actual arrival time. Early arrivals show negative numbers.
CANCELLED --	Cancelled Flight Indicator (1=Yes)
CANCELLATION_CODE --	Specifies The Reason For Cancellation
DIVERTED --	Diverted Flight Indicator (1=Yes)
CRS_ELAPSED_TIME --	CRS Elapsed Time of Flight, in Minutes
ACTUAL_ELAPSED_TIME --	Elapsed Time of Flight, in Minutes
AIR_TIME -- Elapsed Time in Air, in Minutes
DISTANCE -- Distance Traveled, in Miles
CARRIER_DELAY --	Carrier Delay, in Minutes	Analysis
WEATHER_DELAY --	Weather Delay, in Minutes	Analysis
NAS_DELAY --	National Air System Delay, in Minutes
SECURITY_DELAY --	Security Delay, in Minutes
LATE_AIRCRAFT_DELAY --	Late Aircraft Delay, in Minutes

Load the data

```{r}

file_names <- list.files(path = "./",pattern = ".csv")

delays <- do.call('rbind.fill',lapply(file_names,read.csv))

delays <- as.data.table(delays)
delays <- delays[,which(unlist(lapply(delays, function(x)!all(is.na(x))))),with=F]

delays <- as.data.frame(delays)

delays <- delays[complete.cases(delays), ]

str(delays)
```
TODO -- (Describe SGD and the assumptions of the data)

SGD

(great for exceptionally large datasets)

Assumptions of data:

- Linear relationship
- Multivariate normality
- No or little multicollinearity
- No auto-correlation
- Homoscedasticity

Data Preprocessing

```{r}
cols.num <- c("CRS_DEP_TIME", "DEP_TIME", "DEP_DELAY", "TAXI_OUT", "WHEELS_OFF", "WHEELS_ON", "TAXI_IN", 
              "CRS_ARR_TIME", "ARR_TIME", "ARR_DELAY", "CRS_ELAPSED_TIME", "ACTUAL_ELAPSED_TIME")
delays[cols.num] <- sapply(delays[cols.num],as.numeric)
cols.factor <- names(delays) 
cols.factor <- cols.factor[!(cols.factor %in% cols.num)]
delays[cols.factor] <- sapply(delays[cols.factor],as.factor)
delays$FL_DATE <- as.numeric(format(as.Date(delays$FL_DATE), "%Y%m%d"))

str(delays)
```
Test assumptions of data

```{r}

```

Build model

```{r}

```

Stepwise feature selection

```{r}

```

Model Evaluation

```{r}

```

Conclusion:

